---
title: "ESM 244 Lab 1"
author: "Lory Salazar"
date: "1/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

### Read in SF trees data

```{r}
sf_trees <- read.csv(here("data", "sf_trees", "sf_trees.csv"))
```


### Basic wrangling reminders

Find the top 5 highest observations of trees by legal_status, do some wrangling, make a graph.

```{r}
top_5_status <- sf_trees %>% 
  count(legal_status) %>% # recall this counts observation based on variables; combines group_by, summarize, and n
  drop_na(legal_status) %>% 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # moves tree_count to the first column
  slice_max(tree_count, n = 5) # takes top 5 of tree_count
```

### Make a graph

Make a graph of those top 5 observations by legal status.

```{r}
ggplot(data = top_5_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) + # reorders the graph by tree_count instead of alphabetically
  geom_col() +
  labs(x = "Legal Status", y = "Tree Count", title = "Counts of Trees by Legal Status") +
  coord_flip() +
  theme_minimal()
```

### A few more data wrangling refresher examples

Only want to keep observations (rows) for Blackwood Acacia trees.

```{r}
blackwood_acacia <- sf_trees %>% 
  filter(str_detect(species, "Blackwood Acacia")) %>% # look within the data and filter so that within species variable, the Blackwood Acacia tree is detected; useful if the observation doesn't directly match (like here, where we also have the scientific)
  select(legal_status, date, latitude, longitude)

ggplot(data = blackwood_acacia, aes(x = longitude, y = latitude)) + 
  geom_point() 
# gives us a preview of what the map will look like; doesn't register as actual spatial coordinates, however
```

### tidyr::separate() and unite() functions

Useful for combining or separating columns

```{r}
sf_trees_sep <- sf_trees %>% 
  separate(species, 
           into = c("spp_scientific","spp_common"), 
           sep = "::")
```

Example: tidyr::unite()

```{r}
sf_trees_unit <- sf_trees %>% 
  unite("id_status", 
        tree_id:legal_status,
        sep = "_cool!_")
```

### Make some actual maps of Blackwood Acacia trees in SF

`st_as_sf()` to convert the latitude and longitude to spatial coordinates. 

```{r}
blackwood_acacia_sp <- blackwood_acacia %>% 
  drop_na(longitude, latitude) %>% 
  st_as_sf(coords = c("longitude", "latitude")) # write down the variable name storing longitude and latitude

st_crs(blackwood_acacia_sp) = 4326 # from European Petroleum Reference Group for the coordinate reference system

ggplot(data = blackwood_acacia_sp) +
  geom_sf(color = "darkgreen")
```

Read in SF roads shapefile:

```{r}
sf_map <- read_sf(here("data", "sf_map", "tl_2017_06075_roads.shp"))

st_transform(sf_map, 4326)

ggplot(data = sf_map) +
  geom_sf()
```

Combine Blackwood Acacia tree observations and SF roads map:

```{r}
ggplot() +
  geom_sf(data = sf_map, size = 0.1, color = "darkgray") +
  geom_sf(data = blackwood_acacia_sp, size = 0.5, color = "darkgreen") +
  theme_void()
```

Now an interactive map:

```{r}
tmap_mode("view")

tm_shape(blackwood_acacia_sp) +
  tm_dots()
```




